\chapter{Motivation and Introduction}

\section{Background on Credit Card Fraud Detection}
Credit card fraud is a significant and escalating challenge in the global financial ecosystem. As digital transactions, e-commerce platforms, and contactless payments continue to grow rapidly, malicious actors continuously develop sophisticated methodologies to bypass traditional security measures. Fraudulent typologies have evolved from simple physical card theft to complex vectors including phishing, skimming, account takeover, and card-not-present fraud. 

As detailed in contemporary fraud detection literature, building robust defense mechanisms requires understanding the unique nature of financial data \cite{fraud_handbook}. Fraud detection systems operate in a highly dynamic environment characterized by extreme class imbalance—where legitimate transactions vastly outnumber fraudulent ones—and concept drift, as fraudsters continuously alter their tactics to evade detection. Consequently, financial institutions must rely on highly adaptive systems capable of processing immense volumes of transactional data in milliseconds.

\section{The Machine Learning Black Box Dilemma}
To tackle these complex data challenges, financial institutions overwhelmingly adopt statistical learning algorithms complemented by probabilistic thresholds and monitoring layers. Traditional rule-based systems, which relied on rigid, manually defined thresholds, are increasingly being replaced by dynamic, data-driven approaches. Models such as Gradient Boosting Machines and Deep Neural Networks have become the industry standard due to their ability to process massive volumes of transactional data and identify complex, non-linear patterns.

However, these highly complex models inherently act as opaque "black boxes." While they output a precise probability score indicating the likelihood of fraud, they do not provide visibility into the reasoning behind their decisions. From a business and regulatory perspective, predictive accuracy alone is insufficient. In the highly regulated financial industry, automated decision-making is subject to strict scrutiny. Frameworks such as the General Data Protection Regulation mandate transparency. If a legitimate transaction is falsely declined, the institution must be able to justify the decision. Operating a purely black-box model introduces immense business risk, hinders customer trust, and complicates compliance audits.

\section{Explainable AI and Post-Hoc Analysis}
To resolve the tension between high predictive performance and the absolute business requirement for transparency, this thesis explores the implementation of Explainable Artificial Intelligence. Specifically, the research focuses on post-hoc analysis techniques. Post-hoc explainability methods treat the underlying predictive model as a black box and reverse-engineer the decision logic by observing the inputs and outputs after the training phase is complete.

Two of the most robust post-hoc techniques currently driving the field are Local Interpretable Model-agnostic Explanations (LIME) and SHapley Additive exPlanations (SHAP). Both frameworks possess the critical capability of being applied to any underlying algorithm, thereby transforming opaque probabilities into actionable business insights \cite{lime_original,shap_original}. By utilizing these local surrogate models, practitioners can extract human-readable explanations for individual predictions without sacrificing the predictive power of the underlying algorithms.

\section{Research Questions}
To anchor the analytical narrative, three research questions (RQs) are investigated:
\begin{description}
    \item[RQ1] Which evaluation and decision-theoretic arguments justify using precision--recall and threshold-based operating points when fraud prevalence is orders of magnitude smaller than legitimacy?
    \item[RQ2] How should post-hoc explainability theories (local surrogate approximation and additive feature attributions grounded in cooperative games) inform what fraud analysts may legitimately infer from model outputs?
    \item[RQ3] Under which structural assumptions can supervised tabular learners produce probability estimates compatible with qualitative explanation interfaces, despite inevitable distributional drift in adversarial domains?
\end{description}
The methodological substance of each RQ is developed in Chapter~2 (domain theory) and Chapter~3 (proposed mathematical framing). Empirical artefacts are interpreted as illustrative validation rather than exhaustive production certification.

\section{Thesis Objectives and Contributions}
The dissertation prioritises conceptual clarity and formal reasoning about imbalance, probabilistic thresholds, and post-hoc interpretability guarantees. Contemporary fraud detection mandates both statistical rigour \cite{fraud_handbook} and explanatory accountability \cite{lime_original,shap_original,gdpr_text}; bridging these requires analytical tools borrowed from asymmetric classification theory and explainable artificial intelligence surveys \cite{xai_survey}.

The main contributions emphasise theory-aware methodology:
\begin{enumerate}
    \item A structured exposition of imbalance-aware evaluation semantics and operational threshold selection (\emph{versus} naive accuracy optimisation).
    \item A unified discussion of complementary post-hoc explanation paradigms: local surrogate fidelity (LIME), additive Shapley-style attributions (SHAP), and their interplay with constrained financial monitoring contexts.
    \item A methodological architecture that couples predictive scoring $f_\theta$, explicit decision thresholds, and attributable evidence vectors grounded in reproducible preprocessing contracts.
    \item A succinct proof-of-concept instantiation (documented sparingly later) illustrating how algebraic feature vectors align with reproducible artefacts and auditing metadata.
\end{enumerate}

\section{Document Structure}
The remainder of this dissertation unfolds as follows. Chapter~2 reviews state-of-the-art fraud analytics and Explainable AI, emphasising taxonomy, imbalance metrics, limitative properties of explanations, and regulatory motivation. Chapter~3 formalises the proposed methodological solution: stochastic classification model, discriminative surrogates, attribution layers, and operational constraints independent of incidental implementation choices. Chapter~4 concentrates on reproducible experimentation, metric reporting templates, critique of applicability, while referring only minimally to ancillary software artefacts. Chapter~5 summarises conclusions, limitations, and research outlook.