\chapter{State of the Domain}

\section{Background on Fraud Detection and Imbalanced Domains}
Modern fraud detection systems are designed to accurately distinguish between legitimate customer behavior and fraudulent activities in real-time. As established in comprehensive fraud detection literature, building robust predictive models in this sector requires addressing several domain-specific data complexities \cite{fraud_handbook}. 

The primary mathematical challenge in this domain is extreme class imbalance. Fraudulent transactions typically represent a minute fraction of the overall transaction volume, often less than one percent. If a dataset contains 99.9\% legitimate transactions, a naive classifier that predicts all transactions as legitimate will achieve an accuracy of 99.9\%, completely failing its primary objective. Therefore, effective fraud detection relies heavily on robust feature engineering—incorporating variables such as transaction amount, merchant category, geographical location, time of day, and historical spending frequencies—and requires specialized evaluation metrics such as the Area Under the Precision-Recall Curve (AUPRC).

\section{Evolution of Machine Learning Approaches}
Traditional rule-based systems, which rely on rigid, manually defined thresholds, are increasingly being replaced by dynamic, data-driven approaches. Supervised learning techniques utilize historical datasets with labeled fraud cases to train binary classifiers. Algorithms such as Random Forests and Gradient Boosting Machines are highly effective at capturing non-linear relationships within the data. Unsupervised learning and anomaly detection algorithms are also heavily employed to identify novel, zero-day fraud patterns where historical labels are unavailable. 

While these advanced architectures achieve superior predictive accuracy, their internal mechanisms map input features to a probability of fraud through complex, multi-layered mathematical transformations that operate as opaque "black boxes."

\section{Semantic Taxonomy of Model Interpretability}
Explainability notions are heterogeneous; therefore a vocabulary is adopted following survey-level categorisations \cite{xai_survey}.
\begin{enumerate}
    \item \textbf{Transparency depth:} \emph{ante-hoc} / intrinsic interpretability (models whose structure directly supports auditing) versus \emph{post-hoc} interpretation (explainers applied after training).
    \item \textbf{Scope:} \emph{global} understanding summarises unconditional dependencies between predictors and risk; \emph{local} understanding conditions on a realised transaction $x$ and investigates sensitivity of $\hat{y}$ under controlled perturbations.
    \item \textbf{Audience:} compliance-oriented narratives differ from fidelity-oriented diagnostics demanded by modelling teams; ambiguity between \emph{plausible} summaries and \emph{faithful} reconstructions motivates explicit caveats whenever explanations feed operational decisions \cite{xai_survey}.
\end{enumerate}
Fraud operations primarily seek local, post-hoc evidence per alert; therefore later chapters analyse methods that quantify instance-level contributions while referencing global artefacts only as corroborative context.

\section{The Paradigm of Post-Hoc Analysis}
Among post-hoc families, perturbation-derived surrogates and axiomatic attribution maps dominate tabular deployments. Conceptually these methods analyse the directional derivatives of scoring maps $f_\theta:\mathbb{R}^d\to[0,1]$ induced by stochastic classifiers fitted on historical fraud labels.

These techniques treat---at least nominally---the underlying predictor as inaccessible except through query-oracle evaluations. Inputs are manipulated, neighbourhoods are calibrated, cooperative games over feature coalitions may be enumerated (or approximated), and explanatory vectors summarise emergent curvature of $f_\theta$ restricted to neighbourhoods of empirical interest.

The methodological advantage motivating post-hoc choice in financial tooling is encapsulation: the interpretability façade can nominally traverse model upgrades conditioned on preserving API contracts---subject to explanatory stability constraints discussed later.

\section{Advanced Post-Hoc Interpretability Frameworks}
Two of the most prominent model-agnostic, post-hoc explanation frameworks currently driving the field are LIME and SHAP. Both frameworks can be applied to a broad range of predictive models and are widely adopted in explainable AI practice \cite{lime_original,shap_original,xai_survey}.

\subsection{Local Model-Agnostic Explanations (LIME)}
LIME is a post-hoc technique that approximates the predictions of any underlying black-box model by training an interpretable model locally around the prediction of interest. The algorithm operates on the assumption that while a complex model's decision boundary is highly non-linear globally, it is likely linear on a very local scale.

LIME perturbs the input data points to generate a localized dataset, records the black-box model's responses to these variations, and builds a weighted linear model based on the proximity to the original instance. The explanation is generated by minimizing the following objective function:
\begin{equation}
\xi(x) = \underset{g \in G}{\operatorname{argmin}} \mathcal{L}(f, g, \pi_x) + \Omega(g)
\end{equation}
Where $\mathcal{L}$ is a loss function measuring how unfaithful the surrogate model $g$ is in approximating the black box $f$ in the locality defined by $\pi_x$, and $\Omega(g)$ is a complexity penalty ensuring $g$ remains interpretable to human investigators.

\subsection{SHapley Additive exPlanations (SHAP)}
While LIME focuses on local approximation, SHAP provides a unified measure of feature importance based on cooperative game theory. It treats the features of a transaction as "players" in a game where the "payout" is the final fraud probability score. 

SHAP calculates the exact marginal contribution of each feature across all possible feature combinations. The Shapley value $\phi_i$ for a feature $i$ is mathematically defined as:
\begin{equation}
\phi_i = \sum_{S \subseteq N \setminus \{i\}} \frac{|S|! (|N| - |S| - 1)!}{|N|!} \left[ f_x(S \cup \{i\}) - f_x(S) \right]
\end{equation}
Where $N$ is the set of all features, and $S$ is a subset of features not containing $i$. This guarantees that the sum of the individual feature attributions perfectly equals the difference between the model's base expectation and the final prediction, ensuring consistent and mathematically sound explanations for every flagged transaction.

\section{Measuring Classification Quality under Rare Positives}
Class prevalence $\pi=P(y{=}1)$ in retail card streams is microscopic; thus the contingency matrix induces skewed ROC dominance without exposing precision collapse. Accuracy $\frac{TN+TP}{N}$ inflates mechanically when negatives dominate even if $TP\to0$.

The precision--recall curve traces $\text{Recall}(\tau)$ against $\text{Precision}(\tau)$ as the score threshold varies. Integrated measures such as average precision quantify expected precision across recall operating points without assuming symmetry of errors:
\begin{equation}
\text{AP} = \sum_{k} (R_k-R_{k-1}) P_k ,
\end{equation}
where ordered thresholds induce precision $P_k$ at recall breakpoints $R_k$---a pragmatic scalar aligned with asymmetric costs \cite{fraud_handbook}. ROC-AUC still measures ranking separability among positives versus negatives yet can remain high despite economically unacceptable false-positive tides; therefore both ROC and precision--recall summaries deserve parallel reporting.

Operational decisions hinge on thresholds $\tau$ mapping scores to escalate/stop workflows. Equivalent viewpoints cast $\tau$-selection as constrained optimisation balancing expected fraud loss versus investigation budget: denote by $c_{FP}$, $c_{FN}$ asymmetric costs attached to erroneous legitimate vs fraudulent classifications; heuristic Bayes-risk style criteria motivate studying iso-cost surfaces in ROC or PR geometries even when plug-in probabilities are imperfectly calibrated \cite{fraud_handbook}.

\section{Faithfulness, Stability, and Adversarial Drift}
Local explanations hinge on neighbourhoods; minor distributional perturbations---including adversarial adaptations by fraud actors---alter attributions unpredictably unless regularity conditions (Lipschitz moderation, calibrated scores) partially hold \cite{xai_survey}. Therefore explanations must never be anthropomorphised as causal attributions independent of latent confounders in observational transactional data.

From a longitudinal perspective concept drift induces time-varying $f_{\theta}$, rendering static explanation templates stale. Periodic recalibration, monitoring of population stability indices, and revalidation of explanatory alignment with investigator heuristics are theoretical obligations distinct from episodic modelling accuracy gains.

\section{Explainability in Regulated Financial Contexts}
Financial fraud detection is not only a predictive analytics problem, but also a governance problem. In production settings, model outputs may influence high-impact decisions such as transaction blocking, temporary account restrictions, and escalation to manual investigations. As a consequence, institutions must balance statistical performance with accountability, traceability, and decision transparency.

The General Data Protection Regulation (GDPR) and related governance frameworks motivate stronger controls around automated decision support, especially when decisions can materially affect users \cite{gdpr_text}. In practical terms, this means fraud detection platforms should retain model versions, preserve decision logs, and provide interpretable justification artifacts that can be reviewed by analysts and auditors.

\section{Research Gap and Technical Implications}
The literature shows consistent progress in high-performing fraud classifiers, but there remains a practical gap between offline predictive performance and operationally usable explainability. Three technical tensions are especially relevant for this dissertation:
\begin{enumerate}
    \item \textbf{Imbalance versus interpretability:} highly optimized models can improve recall but may produce explanations that are difficult for non-technical investigators to interpret.
    \item \textbf{Local fidelity versus stability:} local explanation methods can be informative for individual cases, yet explanation outputs may vary under small perturbations of the input distribution.
    \item \textbf{Model quality versus workflow cost:} increasing detection sensitivity can raise false-positive volume, directly affecting analyst workload and review latency.
\end{enumerate}

Addressing these tensions requires a design that treats explainability as an integral system requirement rather than a post-deployment add-on. This perspective motivates the architecture proposed in the following chapter, where prediction, explanation, and analyst interaction are developed as a unified workflow.